---
title: "DePaul University Financial Analysis"
subtitle: "Comparative Analysis with Peer Institutions"
format: 
  dashboard:
    theme: cosmo
    orientation: rows
    expandable: true
    scrolling: false
    embed-resources: true
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| echo: false
#| warning: false
#| message: false

library(dplyr)
library(readr)
library(tidyr)
library(ggplot2)
library(stringr)
library(plotly)

# Load data
peer_wide <- read_csv("depaul_peer_wide_metrics.csv", show_col_types = FALSE)

# Ensure correct column types
peer_wide <- peer_wide %>%
  mutate(
    Type = factor(Type),
    School = as.character(School)
  )

# Verify School column exists
if (!"School" %in% names(peer_wide)) {
  stop("Expected a 'School' column in depaul_peer_wide_metrics.csv")
}

# Split DePaul vs peers
depaul <- peer_wide %>% filter(School == "DePaul University")
peers  <- peer_wide %>% filter(School != "DePaul University")
```

# Revenue Analysis

## Row {height="10%"}

DePaul University's revenue structure reveals how the institution finances its operations compared to peer universities. This analysis examines five core revenue sources: tuition and fees, government grants, private gifts, investment returns, and other revenue. Understanding these revenue patterns helps identify DePaul's financial dependencies and diversification relative to similar institutions.

## Row {height="60%"}

### Column

```{r}
#| title: "Revenue Mix: Percentage of Core Revenue"
#| echo: false

# Revenue mix data preparation
rev_vars <- c(
  "Tuition_pct_rev",
  "GovGrants_pct_rev",
  "PrivateGifts_pct_rev",
  "InvReturn_pct_rev",
  "Other_pct_rev"
)

rev_labels <- c(
  Tuition_pct_rev       = "Tuition & fees",
  GovGrants_pct_rev     = "Government grants",
  PrivateGifts_pct_rev  = "Private gifts",
  InvReturn_pct_rev     = "Investment return",
  Other_pct_rev         = "Other core revenue"
)

# Peer averages
peer_rev_avg <- peers %>%
  select(School, Type, all_of(rev_vars)) %>%
  pivot_longer(cols = all_of(rev_vars),
               names_to = "Metric",
               values_to = "Value") %>%
  group_by(Metric) %>%
  summarise(Peer_Avg = mean(Value, na.rm = TRUE), .groups = "drop")

# DePaul
depaul_rev <- depaul %>%
  select(School, all_of(rev_vars)) %>%
  pivot_longer(cols = all_of(rev_vars),
               names_to = "Metric",
               values_to = "DePaul_Value")

# Join
rev_compare <- depaul_rev %>%
  left_join(peer_rev_avg, by = "Metric") %>%
  mutate(Metric_Label = recode(Metric, !!!rev_labels))

# Create plot data
rev_plot_data <- rev_compare %>%
  pivot_longer(cols = c("DePaul_Value", "Peer_Avg"),
               names_to = "Group",
               values_to = "Value") %>%
  mutate(
    Group = recode(Group,
                   DePaul_Value = "DePaul",
                   Peer_Avg = "Peer Average")
  )

# Create plotly chart
plot_ly(
  rev_plot_data,
  x = ~Value,
  y = ~reorder(Metric_Label, Value),
  color = ~Group,
  colors = c("DePaul" = "#0047AB", "Peer Average" = "#87CEEB"),
  type = "bar",
  orientation = "h",
  hovertemplate = paste(
    "<b>%{y}</b><br>",
    "%{x:.1f}%<br>",
    "<extra></extra>"
  )
) %>%
  layout(
    xaxis = list(title = "Percent of core revenue"),
    yaxis = list(title = ""),
    barmode = "group",
    showlegend = TRUE,
    legend = list(orientation = "h", y = -0.15)
  )
```

### Column

```{r}
#| title: "Revenue Per Student (FTE)"
#| echo: false

# Revenue per FTE data preparation
revft_vars <- c(
  "Tuition_rev_perFTE",
  "GovGrants_rev_perFTE",
  "PrivateGifts_rev_perFTE",
  "InvReturn_rev_perFTE",
  "Other_rev_perFTE"
)

revft_labels <- c(
  Tuition_rev_perFTE      = "Tuition & fees",
  GovGrants_rev_perFTE    = "Gov grants",
  PrivateGifts_rev_perFTE = "Private gifts",
  InvReturn_rev_perFTE    = "Investment return",
  Other_rev_perFTE        = "Other core revenue"
)

peer_revft_avg <- peers %>%
  select(School, Type, all_of(revft_vars)) %>%
  pivot_longer(cols = all_of(revft_vars),
               names_to = "Metric",
               values_to = "Value") %>%
  group_by(Metric) %>%
  summarise(Peer_Avg = mean(Value, na.rm = TRUE), .groups = "drop")

depaul_revft <- depaul %>%
  select(School, all_of(revft_vars)) %>%
  pivot_longer(cols = all_of(revft_vars),
               names_to = "Metric",
               values_to = "DePaul_Value")

revft_compare <- depaul_revft %>%
  left_join(peer_revft_avg, by = "Metric") %>%
  mutate(Metric_Label = recode(Metric, !!!revft_labels))

# Create plot data
revft_plot_data <- revft_compare %>%
  pivot_longer(cols = c("DePaul_Value", "Peer_Avg"),
               names_to = "Group",
               values_to = "Value") %>%
  mutate(
    Group = recode(Group,
                   DePaul_Value = "DePaul",
                   Peer_Avg = "Peer Average")
  )

# Create plotly chart
plot_ly(
  revft_plot_data,
  x = ~Value,
  y = ~reorder(Metric_Label, Value),
  color = ~Group,
  colors = c("DePaul" = "#0047AB", "Peer Average" = "#87CEEB"),
  type = "bar",
  orientation = "h",
  hovertemplate = paste(
    "<b>%{y}</b><br>",
    "$%{x:,.0f}<br>",
    "<extra></extra>"
  )
) %>%
  layout(
    xaxis = list(title = "Core revenues per FTE (USD)"),
    yaxis = list(title = ""),
    barmode = "group",
    showlegend = TRUE,
    legend = list(orientation = "h", y = -0.15)
  )
```

## Row {height="30%"}

### Column

```{r}
#| title: "Tuition Dependency Distribution"
#| echo: false

df <- peer_wide %>%
  transmute(
    School = School,
    Tuition_pct_rev = as.numeric(Tuition_pct_rev),
    IsDePaul = ifelse(School == "DePaul University", "DePaul", "Peer")
  ) %>%
  filter(is.finite(Tuition_pct_rev))

plot_ly(df, x = ~Tuition_pct_rev, color = ~IsDePaul,
        colors = c("DePaul" = "#0047AB", "Peer" = "#87CEEB"),
        type = "histogram", 
        nbinsx = 15,
        hovertemplate = paste(
          "Range: %{x}<br>",
          "Count: %{y}<br>",
          "<extra></extra>"
        )) %>%
  layout(
    xaxis = list(title = "Tuition as % of core revenue"),
    yaxis = list(title = "Number of institutions"),
    barmode = "overlay",
    showlegend = TRUE
  )
```

### Column

```{r}
#| title: "Investment Return Distribution"
#| echo: false

df <- peer_wide %>%
  transmute(
    School = School,
    InvReturn_pct_rev = as.numeric(InvReturn_pct_rev),
    IsDePaul = ifelse(School == "DePaul University", "DePaul", "Peer")
  ) %>%
  filter(is.finite(InvReturn_pct_rev))

plot_ly(df, x = ~InvReturn_pct_rev, color = ~IsDePaul,
        colors = c("DePaul" = "#0047AB", "Peer" = "#87CEEB"),
        type = "histogram", 
        nbinsx = 15,
        hovertemplate = paste(
          "Range: %{x}<br>",
          "Count: %{y}<br>",
          "<extra></extra>"
        )) %>%
  layout(
    xaxis = list(title = "Investment return as % of core revenue"),
    yaxis = list(title = "Number of institutions"),
    barmode = "overlay",
    showlegend = TRUE
  )
```

# Expense Analysis

## Row {height="10%"}

How institutions allocate their budgets reflects their priorities and operational models. This section examines expense patterns across seven categories: instruction, research, public service, academic support, student services, institutional support (administration), and other expenses. Comparing DePaul's spending patterns with peer institutions reveals differences in mission focus and resource allocation.

## Row {height="60%"}

### Column

```{r}
#| title: "Expense Mix: Percentage of Core Expenses"
#| echo: false

# Expense mix data preparation
exp_vars <- c(
  "Instruction_pct_exp",
  "Research_pct_exp",
  "PublicService_pct_exp",
  "AcadSupport_pct_exp",
  "StudServices_pct_exp",
  "InstSupport_pct_exp",
  "Other_pct_exp"
)

exp_labels <- c(
  Instruction_pct_exp   = "Instruction",
  Research_pct_exp      = "Research",
  PublicService_pct_exp = "Public service",
  AcadSupport_pct_exp   = "Academic support",
  StudServices_pct_exp  = "Student services",
  InstSupport_pct_exp   = "Institutional support (admin)",
  Other_pct_exp         = "Other core expenses"
)

peer_exp_avg <- peers %>%
  select(School, Type, all_of(exp_vars)) %>%
  pivot_longer(cols = all_of(exp_vars),
               names_to = "Metric",
               values_to = "Value") %>%
  group_by(Metric) %>%
  summarise(Peer_Avg = mean(Value, na.rm = TRUE), .groups = "drop")

depaul_exp <- depaul %>%
  select(School, all_of(exp_vars)) %>%
  pivot_longer(cols = all_of(exp_vars),
               names_to = "Metric",
               values_to = "DePaul_Value")

exp_compare <- depaul_exp %>%
  left_join(peer_exp_avg, by = "Metric") %>%
  mutate(Metric_Label = recode(Metric, !!!exp_labels))

# Create plot data
exp_plot_data <- exp_compare %>%
  pivot_longer(cols = c("DePaul_Value", "Peer_Avg"),
               names_to = "Group",
               values_to = "Value") %>%
  mutate(
    Group = recode(Group,
                   DePaul_Value = "DePaul",
                   Peer_Avg = "Peer Average")
  )

# Create plotly chart
plot_ly(
  exp_plot_data,
  x = ~Value,
  y = ~reorder(Metric_Label, Value),
  color = ~Group,
  colors = c("DePaul" = "#C8102E", "Peer Average" = "#FFB6C1"),
  type = "bar",
  orientation = "h",
  hovertemplate = paste(
    "<b>%{y}</b><br>",
    "%{x:.1f}%<br>",
    "<extra></extra>"
  )
) %>%
  layout(
    xaxis = list(title = "Percent of core expenses"),
    yaxis = list(title = ""),
    barmode = "group",
    showlegend = TRUE,
    legend = list(orientation = "h", y = -0.15)
  )
```

### Column

```{r}
#| title: "Expenses Per Student (FTE)"
#| echo: false

# Expense per FTE data preparation
expft_vars <- c(
  "Instruction_exp_perFTE",
  "Research_exp_perFTE",
  "PublicService_exp_perFTE",
  "AcadSupport_exp_perFTE",
  "StudServices_exp_perFTE",
  "InstSupport_exp_perFTE",
  "Other_exp_perFTE"
)

expft_labels <- c(
  Instruction_exp_perFTE   = "Instruction",
  Research_exp_perFTE      = "Research",
  PublicService_exp_perFTE = "Public service",
  AcadSupport_exp_perFTE   = "Academic support",
  StudServices_exp_perFTE  = "Student services",
  InstSupport_exp_perFTE   = "Institutional support (admin)",
  Other_exp_perFTE         = "Other core expenses"
)

peer_expft_avg <- peers %>%
  select(School, Type, all_of(expft_vars)) %>%
  pivot_longer(cols = all_of(expft_vars),
               names_to = "Metric",
               values_to = "Value") %>%
  group_by(Metric) %>%
  summarise(Peer_Avg = mean(Value, na.rm = TRUE), .groups = "drop")

depaul_expft <- depaul %>%
  select(School, all_of(expft_vars)) %>%
  pivot_longer(cols = all_of(expft_vars),
               names_to = "Metric",
               values_to = "DePaul_Value")

expft_compare <- depaul_expft %>%
  left_join(peer_expft_avg, by = "Metric") %>%
  mutate(Metric_Label = recode(Metric, !!!expft_labels))

# Create plot data
expft_plot_data <- expft_compare %>%
  pivot_longer(cols = c("DePaul_Value", "Peer_Avg"),
               names_to = "Group",
               values_to = "Value") %>%
  mutate(
    Group = recode(Group,
                   DePaul_Value = "DePaul",
                   Peer_Avg = "Peer Average")
  )

# Create plotly chart
plot_ly(
  expft_plot_data,
  x = ~Value,
  y = ~reorder(Metric_Label, Value),
  color = ~Group,
  colors = c("DePaul" = "#C8102E", "Peer Average" = "#FFB6C1"),
  type = "bar",
  orientation = "h",
  hovertemplate = paste(
    "<b>%{y}</b><br>",
    "$%{x:,.0f}<br>",
    "<extra></extra>"
  )
) %>%
  layout(
    xaxis = list(title = "Core expenses per FTE (USD)"),
    yaxis = list(title = ""),
    barmode = "group",
    showlegend = TRUE,
    legend = list(orientation = "h", y = -0.15)
  )
```

## Row {height="30%"}

### Column

```{r}
#| title: "Instructional Spending Distribution"
#| echo: false

df <- peer_wide %>%
  transmute(
    School = School,
    Instruction_pct_exp = as.numeric(Instruction_pct_exp),
    IsDePaul = ifelse(School == "DePaul University", "DePaul", "Peer")
  ) %>%
  filter(is.finite(Instruction_pct_exp))

plot_ly(df, x = ~Instruction_pct_exp, color = ~IsDePaul,
        colors = c("DePaul" = "#C8102E", "Peer" = "#FFB6C1"),
        type = "histogram", 
        nbinsx = 15,
        hovertemplate = paste(
          "Range: %{x}<br>",
          "Count: %{y}<br>",
          "<extra></extra>"
        )) %>%
  layout(
    xaxis = list(title = "Instruction as % of core expenses"),
    yaxis = list(title = "Number of institutions"),
    barmode = "overlay",
    showlegend = TRUE
  )
```

### Column

```{r}
#| title: "Administrative Spending Distribution"
#| echo: false

df <- peer_wide %>%
  transmute(
    School = School,
    InstSupport_pct_exp = as.numeric(InstSupport_pct_exp),
    IsDePaul = ifelse(School == "DePaul University", "DePaul", "Peer")
  ) %>%
  filter(is.finite(InstSupport_pct_exp))

plot_ly(df, x = ~InstSupport_pct_exp, color = ~IsDePaul,
        colors = c("DePaul" = "#C8102E", "Peer" = "#FFB6C1"),
        type = "histogram", 
        nbinsx = 15,
        hovertemplate = paste(
          "Range: %{x}<br>",
          "Count: %{y}<br>",
          "<extra></extra>"
        )) %>%
  layout(
    xaxis = list(title = "Institutional support as % of core expenses"),
    yaxis = list(title = "Number of institutions"),
    barmode = "overlay",
    showlegend = TRUE
  )
```

# Key Metrics

## Row {height="10%"}

This section highlights the most significant differences between DePaul and its peer institutions across both revenue sources and expense categories. These headline metrics provide a snapshot of DePaul's financial position and operational priorities in comparison to similar universities.

## Row {height="45%"}

### Column

```{r}
#| title: "Revenue Comparison: Key Metrics"
#| echo: false

# Headline revenue metrics
headline_rev_vars <- c(
  "Tuition_pct_rev",
  "InvReturn_pct_rev",
  "PrivateGifts_pct_rev",
  "GovGrants_pct_rev"
)

headline_rev_labels <- c(
  Tuition_pct_rev       = "Tuition % of revenue",
  InvReturn_pct_rev     = "Investment return % of revenue",
  PrivateGifts_pct_rev  = "Private gifts % of revenue",
  GovGrants_pct_rev     = "Gov grants % of revenue"
)

depaul_headline_rev <- depaul %>%
  select(School, all_of(headline_rev_vars)) %>%
  pivot_longer(cols = all_of(headline_rev_vars),
               names_to = "Metric",
               values_to = "DePaul")

peer_headline_rev <- peers %>%
  select(School, all_of(headline_rev_vars)) %>%
  pivot_longer(cols = all_of(headline_rev_vars),
               names_to = "Metric",
               values_to = "Value") %>%
  group_by(Metric) %>%
  summarise(Peer_Avg = mean(Value, na.rm = TRUE), .groups = "drop")

headline_rev_compare <- depaul_headline_rev %>%
  left_join(peer_headline_rev, by = "Metric") %>%
  mutate(
    Metric_Label = recode(Metric, !!!headline_rev_labels),
    Difference = DePaul - Peer_Avg
  )

# Create comparison table
plot_ly(
  headline_rev_compare,
  x = ~Difference,
  y = ~reorder(Metric_Label, abs(Difference)),
  type = "bar",
  orientation = "h",
  marker = list(
    color = ~ifelse(Difference > 0, "#0047AB", "#C8102E")
  ),
  hovertemplate = paste(
    "<b>%{y}</b><br>",
    "DePaul: %{customdata[0]:.1f}%<br>",
    "Peer Avg: %{customdata[1]:.1f}%<br>",
    "Difference: %{x:.1f}pp<br>",
    "<extra></extra>"
  ),
  customdata = ~cbind(DePaul, Peer_Avg)
) %>%
  layout(
    xaxis = list(title = "Percentage point difference (DePaul - Peer Avg)"),
    yaxis = list(title = ""),
    showlegend = FALSE
  )
```

### Column

```{r}
#| title: "Expense Comparison: Key Metrics"
#| echo: false

# Headline expense metrics
headline_exp_vars <- c(
  "Instruction_pct_exp",
  "InstSupport_pct_exp",
  "Research_pct_exp",
  "StudServices_pct_exp"
)

headline_exp_labels <- c(
  Instruction_pct_exp   = "Instruction % of expenses",
  InstSupport_pct_exp   = "Institutional support % of expenses",
  Research_pct_exp      = "Research % of expenses",
  StudServices_pct_exp  = "Student services % of expenses"
)

depaul_headline_exp <- depaul %>%
  select(School, all_of(headline_exp_vars)) %>%
  pivot_longer(cols = all_of(headline_exp_vars),
               names_to = "Metric",
               values_to = "DePaul")

peer_headline_exp <- peers %>%
  select(School, all_of(headline_exp_vars)) %>%
  pivot_longer(cols = all_of(headline_exp_vars),
               names_to = "Metric",
               values_to = "Value") %>%
  group_by(Metric) %>%
  summarise(Peer_Avg = mean(Value, na.rm = TRUE), .groups = "drop")

headline_exp_compare <- depaul_headline_exp %>%
  left_join(peer_headline_exp, by = "Metric") %>%
  mutate(
    Metric_Label = recode(Metric, !!!headline_exp_labels),
    Difference = DePaul - Peer_Avg
  )

# Create comparison table
plot_ly(
  headline_exp_compare,
  x = ~Difference,
  y = ~reorder(Metric_Label, abs(Difference)),
  type = "bar",
  orientation = "h",
  marker = list(
    color = ~ifelse(Difference > 0, "#0047AB", "#C8102E")
  ),
  hovertemplate = paste(
    "<b>%{y}</b><br>",
    "DePaul: %{customdata[0]:.1f}%<br>",
    "Peer Avg: %{customdata[1]:.1f}%<br>",
    "Difference: %{x:.1f}pp<br>",
    "<extra></extra>"
  ),
  customdata = ~cbind(DePaul, Peer_Avg)
) %>%
  layout(
    xaxis = list(title = "Percentage point difference (DePaul - Peer Avg)"),
    yaxis = list(title = ""),
    showlegend = FALSE
  )
```

## Row {height="45%"}

### Column {.tabset}

```{r}
#| title: "All Peer Institutions: Tuition Dependency"
#| echo: false

df_scatter <- peer_wide %>%
  transmute(
    School = School,
    Tuition_pct_rev = as.numeric(Tuition_pct_rev),
    InvReturn_pct_rev = as.numeric(InvReturn_pct_rev),
    IsDePaul = ifelse(School == "DePaul University", "DePaul", "Peer")
  ) %>%
  filter(is.finite(Tuition_pct_rev) & is.finite(InvReturn_pct_rev))

plot_ly(
  df_scatter,
  x = ~Tuition_pct_rev,
  y = ~InvReturn_pct_rev,
  color = ~IsDePaul,
  colors = c("DePaul" = "#0047AB", "Peer" = "#CCCCCC"),
  type = "scatter",
  mode = "markers",
  marker = list(size = 10),
  text = ~School,
  hovertemplate = paste(
    "<b>%{text}</b><br>",
    "Tuition: %{x:.1f}%<br>",
    "Investment: %{y:.1f}%<br>",
    "<extra></extra>"
  )
) %>%
  layout(
    xaxis = list(title = "Tuition as % of revenue"),
    yaxis = list(title = "Investment return as % of revenue"),
    showlegend = TRUE
  )
```

```{r}
#| title: "All Peer Institutions: Spending Priorities"
#| echo: false

df_scatter2 <- peer_wide %>%
  transmute(
    School = School,
    Instruction_pct_exp = as.numeric(Instruction_pct_exp),
    InstSupport_pct_exp = as.numeric(InstSupport_pct_exp),
    IsDePaul = ifelse(School == "DePaul University", "DePaul", "Peer")
  ) %>%
  filter(is.finite(Instruction_pct_exp) & is.finite(InstSupport_pct_exp))

plot_ly(
  df_scatter2,
  x = ~Instruction_pct_exp,
  y = ~InstSupport_pct_exp,
  color = ~IsDePaul,
  colors = c("DePaul" = "#C8102E", "Peer" = "#CCCCCC"),
  type = "scatter",
  mode = "markers",
  marker = list(size = 10),
  text = ~School,
  hovertemplate = paste(
    "<b>%{text}</b><br>",
    "Instruction: %{x:.1f}%<br>",
    "Admin: %{y:.1f}%<br>",
    "<extra></extra>"
  )
) %>%
  layout(
    xaxis = list(title = "Instruction as % of expenses"),
    yaxis = list(title = "Institutional support as % of expenses"),
    showlegend = TRUE
  )
```

# Summary

## Row {height="100%"}

### Column

This financial analysis compares DePaul University to a cohort of peer institutions across key revenue and expense metrics. The data reveals several important patterns in how DePaul finances its operations and allocates resources compared to similar universities.

**Revenue Structure**

DePaul demonstrates higher reliance on tuition and fees as a percentage of core revenue compared to peer institutions, while showing lower investment returns and private gift income. This revenue concentration suggests DePaul may be more vulnerable to enrollment fluctuations than peers with more diversified funding sources. On a per-student basis, revenue patterns show where DePaul generates more or less income per FTE compared to similar institutions.

**Expense Allocation**

The expense analysis reveals how DePaul prioritizes its budget relative to peers. Differences in instructional spending, research expenditures, and administrative costs indicate distinct operational models and mission emphases. The per-student expense metrics show whether DePaul invests more or less in specific areas compared to peer institutions, which may reflect different strategic priorities or efficiency levels.

**Strategic Implications**

The key metrics dashboard highlights the most significant differences between DePaul and peer averages. Positive differences indicate areas where DePaul allocates proportionally more resources or generates more revenue, while negative differences show areas of lower emphasis or income. The scatter plots reveal clustering patterns among peer institutions and position DePaul within this competitive landscape.

Understanding these financial patterns helps contextualize DePaul's strategic position. Institutions with similar profiles face similar challenges and opportunities, while outliers may indicate unique missions, markets, or operational approaches. The dashboard provides a foundation for deeper questions about financial sustainability, strategic priorities, and competitive positioning within the higher education marketplace.

**Data Considerations**

Financial metrics represent a single snapshot in time and should be interpreted within the broader context of institutional mission, student demographics, geographic location, and strategic goals. Differences from peer averages are neither inherently positive nor negative but rather reflect distinct institutional characteristics and choices. This analysis provides a comparative framework for understanding DePaul's financial position relative to similar universities.
